---
title: "Cimbi36 spline fit of reference-region"
author: "PPS"
date: "20/07/2020"
output: html_document
---

## Setup

```{r setup, echo=F, warning=F,message=F} 

#knitr::opts_chunk$set(echo = FALSE) #'echo = False' hides *all* code chunks below when knitted 
knitr::opts_chunk$set(warning = FALSE) #'warning = F' hides *all* warnings messages below when knitted 
knitr::opts_chunk$set(message = FALSE) #'message = F' hides *all* messages below when knitted 

#set path to the parent folder of the project
knitr::opts_knit$set(root.dir = normalizePath("/Users/pontus.sigray/Documents/NRU/Bootstrapper"))

#load packages (install from CRAN unless otherwise specified)
library(kinfitr) #devtools::install_github('mathesong/kinfitr')
library(tidyverse)
library(knitr)
library(kableExtra)
library(mgcv)
library(ggforce)
library(furrr)
library(ggplot2)
library(magrittr)

#Graphics theme
theme_set(theme_light())


```

## Fetch all data

```{r}

#fetch all ROI TACs and blood files
source('./R/cimbi36_TAC_analysis/Prep/fetch_ROITAC_and_Blood_Data_TrT.R')

path.PET1 <- './RawData/cimbi36_TrT/Test/'
path.PET2 <- './RawData/cimbi36_TrT/Retest/'

PET1.data <- fetch_ROITAC_and_Blood_Data_TrT(path = path.PET1,PETNo = 1)
PET2.data <- fetch_ROITAC_and_Blood_Data_TrT(path = path.PET2,PETNo = 2)

alldat <- dplyr::bind_rows(PET1.data,PET2.data)

#fetch database overview

db.df <- xlsx::read.xlsx(file = './RawData/cimbi36_Database_overview/cimbi36_HC_Database_overview.xlsx',sheetIndex = 1)
db.df <- select(db.df,c(CIMBI.ID,Cimbi.36.ID))

alldat <- dplyr::left_join(alldat,db.df,by = c("Subjname"="Cimbi.36.ID"))

alldat$PET <- paste0(alldat$CIMBI.ID,'_',alldat$PETNo)

alldat <- select(alldat,c(PET,CIMBI.ID,PETNo,Subjname,tacs,blooddata))

```


# Add gaussian noise to cerebellum data
```{r}

add.noise <- function(tacs.inp){
  tacs.inp$Total_cb.noise <- tacs.inp$Total_cb + rnorm(n = length( tacs.inp$Total_cb), mean = 0,sd = 1)
  
  tacs.inp$Total_cb.noise[tacs.inp$Total_cb.noise<0] <- 0
  
  return(tacs.inp)
}


alldat <- alldat %>%
  mutate( tacs = map(.x = tacs,.f = add.noise))


```


# Fit spline to Cerebellum

```{r}


splinefit <- function(tacs){
  
  gam.fit <- mgcv::gam(formula = (Total_cb ~ s(Times , bs = 'cr', k = 14)),
                       sp = 3,
                       data = tacs)
  
  return(gam.fit)
  
}

splinefit.noise <- function(tacs){
  
  gam.fit <- mgcv::gam(formula = (Total_cb.noise ~ s(Times , bs = 'cr', k = 30)),
                       data = tacs)
  
  return(gam.fit)
  
}


alldat <- alldat %>%
  mutate( spline.fit = map(.x = tacs,.f = splinefit),
          spline.fit.noise = map(.x = tacs,.f = splinefit.noise),)

#Add fitted values to tacs
add.spline.fit <- function(tacs.inp,spline.fit.inp){
  
  Total_cb.splinefit <- as.numeric(predict(spline.fit.inp))
  Total_cb.splinefit[1] <- 0 #set first value to zero 
  
  tacs.inp$Total_cb.splinefit <- Total_cb.splinefit
  
  return(tacs.inp)
  
}

#Add fitted values to tacs
add.spline.fit.noise <- function(tacs.inp,spline.fit.inp){
  
  Total_cb.splinefit.noise <- as.numeric(predict(spline.fit.inp))
  Total_cb.splinefit.noise[1] <- 0 #set first value to zero 
  
  tacs.inp$Total_cb.splinefit.noise <- Total_cb.splinefit.noise
  
  return(tacs.inp)
  
}

alldat <- alldat %>%
  mutate( tacs = purrr::map2(tacs,spline.fit, .f = add.spline.fit  ),
          tacs = purrr::map2(tacs,spline.fit.noise, .f = add.spline.fit.noise  ))



```

#Plot cerebellum TACs and spline fits

```{r,eval = F}
plotTACs <- function(tacs, Subjname){
  
  plotTAC <- tacs %>%
  select( c( Times,Total_cb.noise) ) %>% #select the Regions to model  
  ggplot(aes(x=Times, y=Total_cb.noise)) +
    geom_point() +
    geom_line(data = tacs,aes(x = Times,y = Total_cb.splinefit.noise), col = 'red') +
  labs(title=Subjname,
       y="Radioactivity (kBq)",
       x = "Time (min)")
  
}

alldat <- alldat %>%
  mutate(plotTACs = map2(.x = tacs,.y = Subjname, ~plotTACs(tacs = .x,Subjname = .y) ) )

print(alldat$plotTACs)

```

## Create weights

Since sqrt(duration) weights might be very off for alldat I'm going to use the build in generic weights function of kinfitr. 

```{r}

create_weights <- function(tacs){
  
  new_weights <- kinfitr::weights_create(t_start = c(tacs$Times - tacs$Duration/2),
                          t_end = c(tacs$Times + tacs$Duration/2),
                          tac = tacs$Temporal.Cortex,
                          radioisotope = c("C11"),
                          method = 2,
                          minweight = 0.7)
  
  tacs$Weights <- new_weights
  return(tacs)
  
}
  
alldat %<>%
  mutate(tacs = purrr::pmap(.l = list(tacs),.f = create_weights))
```


## Make alldat into even longer nested format by spreading ROIs

```{r}

alldatLong <- alldat %>%
  select(PET,Subjname, PETNo, tacs) %>%
  unnest() %>%
  gather(Region, TAC, -c(PET,Subjname, PETNo, Total_cb,Total_cb.noise, 
                         Total_cb.splinefit,Total_cb.splinefit.noise,
                         Times, Weights)) %>%
  group_by(PET,Subjname, PETNo, Region) %>%
  nest(.key = 'tacs') %>%
  filter(Region != 'Duration')


```

#Fit SRTM

```{r}


alldatLong <- alldatLong %>%
  filter(Region == c('Striatum') |
           Region == c('Temporal.Cortex') |
           Region == c('Frontal.Cortex'))

#Make function for fitting  srtm using raw cer values
srtm.cb <- function(tacs) {
  kinfitr::srtm(t_tac = tacs$Times, 
                reftac = tacs$Total_cb,
                roitac = tacs$TAC, 
                weights = tacs$Weights,
                multstart_iter = 1) 
}

srtm.cb.noise <- function(tacs) {
  kinfitr::srtm(t_tac = tacs$Times, 
                reftac = tacs$Total_cb.noise,
                roitac = tacs$TAC, 
                weights = tacs$Weights,
                multstart_iter = 1) 
}

#Make function for fitting  srtm using smoothed cer values
srtm.cb.spline <- function(tacs) {
  kinfitr::srtm(t_tac = tacs$Times, 
                reftac = tacs$Total_cb.splinefit,
                roitac = tacs$TAC, 
                weights = tacs$Weights,
                multstart_iter = 1) 
}

#Make function for fitting  srtm using smoothed cer values
srtm.cb.spline.noise <- function(tacs) {
  kinfitr::srtm(t_tac = tacs$Times, 
                reftac = tacs$Total_cb.splinefit.noise,
                roitac = tacs$TAC, 
                weights = tacs$Weights,
                multstart_iter = 1) 
}



#Fit 2TCM and calculate VT, VS and BPND from rate constants
alldatLong <- alldatLong %>%  
  mutate(fit_srtm.cb = purrr::map(tacs, srtm.cb)) %>%
  mutate(BPND.cb = map_dbl(fit_srtm.cb, c('par', 'bp'))) %>%
  
  mutate(fit_srtm.cb.noise = purrr::map(tacs, srtm.cb.noise)) %>%
  mutate(BPND.cb.noise = map_dbl(fit_srtm.cb.noise, c('par', 'bp'))) %>%
  
  mutate(fit_srtm.cb.spline = purrr::map(tacs, srtm.cb.spline)) %>%
  mutate(BPND.cb.spline = map_dbl(fit_srtm.cb.spline, c('par', 'bp'))) %>%
  
  mutate(fit_srtm.cb.spline.noise = purrr::map(tacs, srtm.cb.spline.noise)) %>%
  mutate(BPND.cb.spline.noise = map_dbl(fit_srtm.cb.spline.noise, c('par', 'bp'))) 

 

```

## Plot BPND from cerebellum v.s. BPND from cerebellum spline fit 

```{r}

plotData <- select(alldatLong, c(PET, Subjname, PETNo, Region, BPND.cb, BPND.cb.noise, BPND.cb.spline,BPND.cb.spline.noise ))

plotDataNoNoise <- select(plotData,c( PET, Subjname, PETNo, Region,BPND.cb,BPND.cb.spline))
plotDataNoNoiseLong <- gather(plotDataNoNoise, Condition, BPND.No.Noise, BPND.cb:BPND.cb.spline)

plotDataNoNoiseLongRaw <- filter(plotDataNoNoiseLong,Condition == 'raw')

#recode condition
plotDataNoNoiseLong$Condition[plotDataNoNoiseLong$Condition == 'BPND.cb'] <- 'raw'
plotDataNoNoiseLong$Condition[plotDataNoNoiseLong$Condition == 'BPND.cb.spline'] <- 'spline'

plotDataNoise <- select(plotData,c( PET, Subjname, PETNo, Region, BPND.cb.noise, BPND.cb.spline.noise))
plotDataNoiseLong <- gather(plotDataNoise, Condition, BPND.Noise, BPND.cb.noise:BPND.cb.spline.noise  )

plotDataNoiseLongSpline <- filter(plotDataNoiseLong,Condition == 'spline')

#recode condition
plotDataNoiseLong$Condition[plotDataNoiseLong$Condition == 'BPND.cb.noise'] <- 'raw'
plotDataNoiseLong$Condition[plotDataNoiseLong$Condition == 'BPND.cb.spline.noise'] <- 'spline'

plotDataLong <- left_join(plotDataNoNoiseLong,plotDataNoiseLong)


ggplot(data = plotDataLong, aes(x = BPND.No.Noise, y = BPND.Noise, shape = Condition, color = Condition) ) + 
  xlab(label = 'BPND noise free')+ 
  ylab(label = 'BPND noisy')+ 
  geom_abline(slope = 1,intercept = 0) + 
  geom_point()

#ggsave(filename = '/Users/pontus.sigray/Desktop/BPND_raw_vs_spline_CER.png',device = 'png',width = 22.5,height = 15,units = 'cm',dpi = 350)


ggplot(data = plotDataLong, aes(x = BPND.No.Noise, y = BPND.Noise, shape = Condition, color = Condition) ) + 
  xlab(label = 'BPND noise free')+ 
  ylab(label = 'BPND noisy')+ 
  geom_abline(slope = 1,intercept = 0) + 
  ylim(c(0,2.1)) + 
  xlim(c(0,2.1)) + 
  geom_point()



```
## Extra: CV-loo grid search knots and lambda 
```{r, eval = F}

tacs <- alldat$tacs[[3]]

#CV.looc.gridsearch <- function(tacs){
knot.grid <- seq(from = 5, to = 30, by = 1 )
lamda.grid <- 0.0001 * 10^(seq(0,10,length.out = 100) )
RSS.out <- NULL

for (k in 1:length(knot.grid) ){

knots = knot.grid[k]
  
for(i in 1:length(lamda.grid)){
  
  res.loo <- NULL
  for (p in 1:length(tacs$Times)){
    
    loo.point <- tacs$Times[p] 
    tacs.loo <- tacs %>%
      filter(Times != loo.point)
    
    gam.fit <- mgcv::gam(formula = (Total_cb.noise ~ s(Times , bs = 'cr', k = knots)),
                         sp= lamda.grid[i] ,
                           data = tacs.loo)
    
    res.loo[p] <- tacs$Total_cb.noise[p] - predict(gam.fit,newdata = tibble(Times = loo.point) )
    
    
  }
  
  RSS[i] <- sum(res.loo^2)
}

RSS.out <- cbind(RSS.out,RSS)
print(k)
}


rownames(RSS.out) <- paste0('lambda_',lamda.grid)
colnames(RSS.out) <- paste0('knot_',knot.grid)

heatmap(log(RSS.out),  Colv=NA, Rowv=NA, scale='none')
RSS.min.idx <- which(RSS.out == min(RSS.out), arr.ind = TRUE)

out <- list(RSS.min.idx = RSS.min.idx)
#}


sp <- 0.5
knots <-8


gam.fit <- mgcv::gam(formula = (Total_cb.noise ~ s(Times , bs = 'cr', k = knots)),sp = sp,
                     data = tacs)


#lambda <- gam.fit$sp; lambda
#length(gam.fit$smooth[[1]]$xp)
tacs$Total_cb.splinefit.noise <- as.numeric(predict(gam.fit))
tacs$Total_cb.splinefit.noise[1] <- 0
plot(tacs$Times,tacs$Total_cb.noise, xlim = c(0,5), ylim = c(0,14) )
points(tacs$Times,tacs$Total_cb.splinefit.noise,type = 'l')
points(tacs$Times,tacs$Total_cb.splinefit.noise,col = 'blue')

points(tacs$Times,tacs$Total_cb,col = 'red')


abline(v = gam.fit$smooth[[1]]$xp, type = 2)

##fit and compare


spline.fit.noise <- kinfitr::srtm(t_tac = tacs$Times,reftac = tacs$Total_cb.splinefit.noise,roitac = tacs$Frontal.Cortex,weights = tacs$Weights)

raw.fit <- kinfitr::srtm(t_tac = tacs$Times,reftac = tacs$Total_cb,roitac = tacs$Frontal.Cortex,weights = tacs$Weights)

raw.fit.noise <- kinfitr::srtm(t_tac = tacs$Times,reftac = tacs$Total_cb.noise,roitac = tacs$Frontal.Cortex,weights = tacs$Weights)

raw.fit$par$bp
raw.fit.noise$par$bp
spline.fit.noise$par$bp

```

